---
title: "IODS course project"
author: "H. Sanez"
output:
  html_document:
    theme: cosmo
    toc: true
    #toc_depth: 2
    toc_depth: 3 #HS
    toc_float: true #HS
    fig_caption: true
    fig_width: 6
    fig_height: 4
---

***

# Introduction to Open Data Science - Course Project

| Chapter | Topic | Max points | Book Ch (K.V.)| Book Ch R for HDS | Assignment done? |
|:-------:|:------------:|:------:|:------:|:------:|:------:|
| 1 | Start me up! | 20 |1 & 2| - |&cross;|
| 2 | Regression and model validation |5+15| 3 & 4 | 7 |&cross;|
| 3 | Logistic regression | 5+15 | 5 & 6 | 7 |&cross;|
| 4 | Clustering and classification | 15+5 |-|12, 17 & 18|&cross;|
| 5 | Dimensionality reduction techniques | 5 + 15 | 13 & 14 | - |&cross;|
| 6 | Analysis of longitudinal data | 5 + 15 | 8 & 9 | - |&cross;|
| Misc | Misc |||||



```{r child = "chapter1.Rmd"}
# This will include Chapter 1 (that is updated in its own file) in the document.
```
  
***
  
```{r child = "chapter2.Rmd"}
# This will include Chapter 2 (that is updated in its own file) in the document.
```

***

```{r child = "chapter3.Rmd"}
# This will include Chapter 3 (that is updated in its own file) in the document.
```

***

```{r child = "chapter4.Rmd"}
# This will include Chapter 4 (that is updated in its own file) in the document.
```

***

```{r child = "chapter5.Rmd"}
# This will include Chapter 5 (that is updated in its own file) in the document.
```

***

```{r child = "chapter6.Rmd"}
# This will include Chapter 6 (that is updated in its own file) in the document.
```

***

(more chapters to be added similarly as we proceed with the course!)

