---
title: "IODS course project"
author: "H. Sanez"
output:
  html_document:
    theme: cosmo
    toc: true
    #toc_depth: 2
    toc_depth: 3 #HS
    toc_float: true #HS
    fig_caption: true
    fig_width: 6
    fig_height: 4
---

***

# Introduction to Open Data Science - Course Project

| Chapter | Topic | Max points | Book Ch (K.V)| Book Ch R for HDS | Assignment done? |
|:-------:|:------------:|:------:|:------:|:------:|:------:|
| 1 | Start me up! | 20 |1 & 2  | - |&cross;|
| 2 | Regression and model validation |5+15| 3 & 4 |  7 ||
| 3 | Logistic regression |  |  |||
| 4 | Clustering and classification |||||
| 5 | Dimensionality reduction techniques |||||
| 6 | Analysis of longitudinal data |||||
| Misc | Misc |||||


```{r child = "chapter1.Rmd"}
# This will include Chapter 1 (that is updated in its own file) in the document.
```
  
***
  
```{r child = "chapter2.Rmd"}
# This will include Chapter 2 (that is updated in its own file) in the document.
```

***

(more chapters to be added similarly as we proceed with the course!)

